---
title: "Data Cleaning & Explorations"
author: "Abdul-lateef Abiola Hussein"
date: "2022-09-26"
output:
  pdf_document: default
  html_document: default
  word_document: default
---

# Palmer Penguins Cleaning and Exploration

This projects was done to show case my skill in using R programming language for Data cleaning process, exploration and documentation. 

## Setting up suitable working environment
Procedures necessary include 
Installation of palmerpenguins packages which contains the raw data needed for this analysis. Loading the packages needed for data cleaning, data manipulation, visualization such as;
'tidyverse', 'tidyr','dplyr','ggplot','janitor', and 'skimr'.


```{r loading packages}
library(tidyverse)
library(dplyr)
library(tidyr)
library(janitor)
library(stringr)
library(skimr)
library(palmerpenguins)

```

## Prepare and processing Phase.
These phases involves inspecting the 'penguins_raw' data, extraction of the needed variables , data cleaning and data validations.

```{r Data examination }
View(penguins_raw)

colnames(penguins_raw)

skim_without_charts(penguins_raw)
```
 
```{r Data Extractions of the needed variables}
peng <- penguins_raw %>%
  clean_names()

colnames(peng)

peng_data<-peng %>% 
  select(c(species,region,island,culmen_length_mm,
           culmen_depth_mm,flipper_length_mm,body_mass_g,sex))
peng_data

```
 
```{r Data Cleaning Processes}
peng_cleaned<-peng_data %>% 
  drop_na(sex, culmen_length_mm,culmen_depth_mm,flipper_length_mm,body_mass_g)

sep<- peng_cleaned %>% 
  separate(species, into = c("a","b"))

peng_cleaned_data <- sep %>% 
  unite("species", c(a,b), sep = " " )

skim_without_charts(peng_cleaned_data)
```

```{r Data Formatting}
peng_cleaned_data <-peng_cleaned_data %>% 
  mutate(body_mass_kg = body_mass_g/1000, sex= tolower(sex)) %>% 
  select(-body_mass_g)
```

## Analysis Phase
This phase involves manipulating the cleaned penguins datasets in order to extract answers from the heart of the data.

###1.Which specie of penguins is dominant.

```{r}
peng_cleaned_data %>% 
  group_by(species) %>% 
  count(species) %>% 
  mutate(percent_species= n*100/333)

```

```{r}
peng_cleaned_data %>% 
  ggplot(aes(x=species, fill= species)) + geom_bar(show.legend= FALSE)+
  labs(title = "Dorminant Penguins Species", y= "population",
       subtitle = "Adelie Penguin spcies are the largest with 43.8% of the population") 
```

### #2. Which gender dominates the population

```{r}
peng_cleaned_data %>% 
  group_by(sex) %>% 
  count(sex) %>% 
  mutate(percent_sex = n*100/333)

```

```{r}
peng_cleaned_data %>% 
  ggplot() + geom_bar(mapping = aes(x= sex, fill= sex), show.legend = FALSE) + 
  labs(title ="Population By Gender",y= "population", 
       subtitle = "There is 168 males penguins and 165 females penguins in the population.")
```

### 3.What's the correlation between flipper_length and body_mass_kg

```{r}
peng_cleaned_data %>% 
  select(flipper_length_mm,body_mass_kg)
```

```{r}
peng_cleaned_data %>% 
  ggplot(aes(x= body_mass_kg,y =flipper_length_mm)) + geom_point()+
  geom_smooth(method = "loess") +
  labs(title = "Relationship between Flipper Length and Body Mass", subtitle =
         "Positive correlation i.e filpper length increases with the body mass ")
```

```{r}
peng_cleaned_data %>% 
  ggplot(aes(x= body_mass_kg,y =flipper_length_mm, color= species)) + geom_point()+
  geom_smooth(method = "loess") +  facet_wrap(~species) +
  labs(title = "Flipper Length Vs Body Mass Relationship in Each Species",
       subtitle= " The lowest positive correlation is observed in Adellie species")
```

### 4.Which species has the highest average body_mass_kg

```{r}
peng_cleaned_data %>% 
  group_by(species) %>% 
  summarise(avg_body_mass = mean(body_mass_kg)) %>% 
  ggplot(aes(x= species, y= avg_body_mass, fill= species)) + geom_col(show.legend = FALSE) +
  labs(title = "Average Body Mass(kg) Of Each Species",
       subtitle = "Gentoo penguins species has the highest avaerage mass of 5.09kg.
       Adelie and chinstrap penguins has 3.71kg and 3.73kg respectively.")
```


 ### 5.Which island has the highest population of penguins and why?

```{r}
peng_cleaned_data %>% 
  group_by(island) %>% 
  count(island) %>% 
  arrange(-n, .by_group = TRUE)
```
 
```{r}
peng_cleaned_data %>% 
  group_by(island) %>% 
  count(island) %>% 
  arrange(-n, .by_group = TRUE) %>% 
  ggplot(aes(x= island, y= n, fill = island)) + geom_col() + 
  labs(title = "Island with the Highest Population of penguins")
```
 
```{r}
peng_cleaned_data %>% 
  group_by(island,species) %>% 
  count(species) %>% 
  arrange(-n, .by_group = TRUE) %>% 
  ggplot(aes(x= island, y= n, fill = species)) + geom_col() + 
  labs(title = "Island with the Highest Population of penguins",
  caption ="Biscoe island contains all the gentoo penguins (119) used in this dataset and 30% (44) of total adelie speceis.
  Dream island contains all Chinstrap penguin (68) and 38% i.e(55) of gentoo penguin.
  Torgersen island contains only gentoo penguins which is 32%, i.e (47), of the total gentoo population in this dataset")
```
 
 ### 6. What is the correlation between the body_mass_kg and culmen_length_mm in general
 
```{r}
peng_cleaned_data %>% 
  select(culmen_length_mm,body_mass_kg)

```

```{r}
peng_cleaned_data %>% 
  ggplot(aes(x=body_mass_kg,y=culmen_length_mm)) + geom_point() +
  geom_smooth(method = "loess") +
  labs(title = "Relation Between Body mass and culmen length")
```

 ### 6b. What is the correlation between the body_mass_kg and culmen_length_mm (faceting by species)

```{r}
peng_cleaned_data %>% 
  ggplot(aes(x=body_mass_kg,y=culmen_length_mm, line = species, shape= species, color=species)) + 
  geom_jitter() + geom_smooth(method = "loess") +facet_wrap(~species) +
  labs(title= "Relation Between Body mass and culmen length in each species",
       subtitle ="The two variables show strong positive relation in Chinstrap penguins.
       But less strong positive relationship in the other two species")
```
 
 ### 7. What is the correlation between the culmen_depth_mm and culmen_length_mm in general

```{r}
peng_cleaned_data %>% 
  select(culmen_length_mm,culmen_depth_mm)
```
 
```{r}
peng_cleaned_data %>% 
  ggplot(aes(x=culmen_length_mm,y=culmen_depth_mm)) + 
  geom_jitter() + geom_smooth(method = "loess") +
  labs(title = "Relation between Culmen Depth and Culmen Length",
       subtitle = "There is no any any relationship between the two variable.Hence,the relationship is non-linear")
```
 
#### Thanks :)